## Data validation {#sec-validate}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))
```


```{r, include = FALSE}
library(dplyr)
library(assertr)

set.seed(123)
n <- 20

# Step 1: create dataset
female_egg_data <- tibble(
  female_id = 1:n,
  treatment = rep(c("A","B"), times =10),
  age_days = sample(1:30, n, replace = TRUE),
  eggs_laid = rpois(n, lambda = 50),
  eggs_hatched = pmin(eggs_laid, rpois(n, lambda = 45))
)

# Step 2: Introduce some deliberate issues
female_egg_data$age_days[1] <- 0                     # invalid age
female_egg_data$eggs_laid[2] <- 120 # extreme eggs_laid
female_egg_data$eggs_hatched[3] <- female_egg_data$eggs_laid[3] + 5  # eggs_hatched > eggs_laid

female_egg_data

female_egg_data <- female_egg_data |> mutate(across(where(is.numeric),
                                 .fns = as.integer))

write_csv(female_egg_data, here::here("files", "female_egg_data.csv"))


```

## Learning Objectives:

This section covers different methods for validating the assumptions about our data in R, by the end of this chapter you will be able to:

- Explain why data validation matters for reproducible research.

- Use `assertr` functions (`assertr::verify()` , `asserrr::assert()`, `assertr::insist()`) to check data.

- Write automated checks for numeric ranges and categorical values.

- Combine validation checks into a simple, readable pipeline.

- Interpret errors and handle validation failures gracefully.


## Why validate data

Before we analyse data we need to trust it

Think about the assumptions we build into or data:

- Counts are always positive

- Column B is always < than Column A

- Measurements are within acceptable ranges

> Remember Garbage in = Garbage out


## Our dataset

Here’s our experimental dataset: each row is a female insect, with basic life-cycle and reproduction data

```{r, eval = TRUE, echo = FALSE}
downloadthis::download_link(
  link = "https://raw.githubusercontent.com/Philip-Leftwich/Oct-Intro-Analytics/refs/heads/dev/files/female_egg_data.csv",
  button_label = "Download female egg data as csv",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```

```{r}

female_egg_data

```

:::{.task}
::::{.task-header}
Question
::::
::::{.task-container}

If this were your dataset, what checks would you want to include?


`r hide("Solution")`

| Column         | What we want to ensure               | assertr check                                                                      |
| -------------- | ------------------------------------ | ---------------------------------------------------------------------------------- |
| `female_id`    | No missing, unique, positive integer | `assert(not_na, female_id)`, `assert(is_uniq, female_id)`                          |
| `treatment`    | Only “A” or “B”                      | `verify(treatment %in% c("A","B"))`                                                |
| `age_days`     | >0 and reasonable range              | `assert(within_bounds(1, 30), age_days)`                                           |
| `eggs_laid`    | >0, within 3 SDs                     | `assert(within_bounds(1, Inf), eggs_laid)`, `insist(within_n_sds(3), eggs_laid)`   |
| `eggs_hatched` | ≥0 and ≤ eggs_laid                   | `assert(within_bounds(0, Inf), eggs_hatched)`, `verify(eggs_hatched <= eggs_laid)` |


`r unhide()`

::::
:::

## Run an assertion check

With `assertr` by default if a validation check fails the code will throw an error: 

```{r, eval = FALSE}
library(assertr)

female_egg_data |>
  verify(age_days > 0)                      # hatched ≤ laid


```


## assert, insist, verify

The assertr package (@R-assertr) has three main functions:

- `verify()` - **is this statement true for the whole data?**

```{r, eval = F}
df |>  verify(has_all_names("female_id","treatment","age_days","eggs_laid","eggs_hatched"))

```

- `assert()` - **applies a prediction row-by-row to the specified columns**

```{r, eval = F}
df |>  assert(is.numeric, female_id, age_days, eggs_laid, eggs_hatched)
```

- `insist()` - **Flag values using a rule that depends on the whole column**


```{r, eval = F}
df |>  insist(within_n_sds(3), eggs_laid)

```


### error functions

The `error_fun` argument defines what happens when a validation check produces an error. By default it prints a summary of the errors and halts the code. 

This may not be desirable if we wish to run a longer pipeline or many error checks, there are multiple options including `just_warn` script execution continues and a warning is printed. 

```{r}
female_egg_data |>
  verify(age_days > 0,
         error_fun = just_warn)  

```

## Pipelines

We can string together multiple checks into a single pipeline, we don't *need* `assertr`, it can be achieved using `dplyr` but this package makes pipelines easier. 

:::{.panel-tabset}
# Assertr
```{r}
data_checks <- function(female_egg_data, just_warn, is.numeric, female_id, age_days, eggs_laid, eggs_hatched, is.character, treatment, is_uniq, not_na) {
  female_egg_data  |> 
      verify(has_all_names("female_id", "treatment",
                           "age_days", "eggs_laid", "eggs_hatched"),
             error_fun = just_warn) |>
      assert(is.numeric, female_id, age_days, eggs_laid, eggs_hatched,
             error_fun = just_warn) |>
      assert(is.character, treatment, error_fun = just_warn) |>
      assert(is_uniq, female_id, error_fun = just_warn) |>
      assert(not_na, female_id, treatment, age_days,
             eggs_laid, eggs_hatched, error_fun = just_warn) |>
      assert(in_set("A", "B"), treatment) |> 
      assert(within_bounds(1, 30), age_days, error_fun = just_warn) |>
      verify(eggs_hatched <= eggs_laid, error_fun = just_warn) |>
      insist(within_n_sds(3), eggs_laid, error_fun = just_warn)
}

```


# dplyr


```{r}
# Step 1: Get summary stats for eggs_laid
female_egg_data_summary <- female_egg_data |> 
  summarise(
    mean_eggs_laid = mean(eggs_laid, na.rm = TRUE),
    sd_eggs_laid   = sd(eggs_laid, na.rm = TRUE)
  )

# Check column classes
female_egg_data |> 
  summarise(across(everything(), ~ class(.)[1])) %>%
  tidyr::pivot_longer(everything(),
                      names_to = "column",
                      values_to = "type")

female_egg_data |> 
  filter(duplicated(female_id))

female_egg_data_report <- female_egg_data |>
  mutate(
    age_invalid           = age_days <= 0,
    eggs_laid_invalid     = eggs_laid <= 0,
    eggs_laid_outlier     = eggs_laid < (female_egg_data_summary$mean_eggs_laid - 3*female_egg_data_summary$sd_eggs_laid) |
                             eggs_laid > (female_egg_data_summary$mean_eggs_laid + 3*female_egg_data_summary$sd_eggs_laid),
    eggs_hatched_invalid  = eggs_hatched > eggs_laid
  )

# Step 3: view rows with any issues filter == TRUE
female_egg_data_report |> 
  filter(age_invalid | eggs_laid_invalid | eggs_laid_outlier | eggs_hatched_invalid)

```

:::


## Options for within group datachecks

`assertr` is designed for whole dataset validation, but it can make sense for some `assertr::insist()` rules that these are applied in a group specific way

:::{.task}
::::{.task-header}
Question
::::
::::{.task-container}

When might we want to consider groups for some data validations?

`r hide("Solution")`

For data within the `insist` function we compare data across the entire column - in these examples we might check data is within certain deviations of the mean. But if there are multiple groups, this makes sense to perform within groups.

```{r}
female_egg_data |> 
  group_by(treatment) |> 
dplyr::group_modify(~ .x |> insist(within_n_sds(3), eggs_laid, error_fun = just_warn))

```

`r unhide()`

::::
:::



## Summary

You now know how to:

- Use `assertr` to check data integrity before analysis.

- Validate numeric and categorical variables.

- Combine checks into a tidy pipeline.

- Control what happens when checks fail.

Automated validation transforms your scripts into self-auditing workflows — they document and test your assumptions each time the data changes.


