## Function based data checks {#sec-validate-functional}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
```


```{r, include = FALSE}
library(dplyr)
library(assertr)

female_egg_data <- read_csv(here::here("files", "female_egg_data.csv"))


```

## Motivation

We have run through the principles of organised data dictionaries and data validation. Once we understand the consistent types of data(sets) we encounter it is possible to start building data validation **pipelines**.

- Pipelines are collections of custom data functions that can apply consistent data cleaning steps to different data that fits a particular layout

- Pipelines can help speed up the process of data validation, analysis and plotting. 

### Examples

- Parameterised reports - simple summary reports that can be generated from a dataset with a button push

- Parameterised data import and validation pipelines - our second example and presented here:


## The data

In the previous chapters we worked through the process of applying data validation checks to a dataset on female age, fertility and fecundity according to treatments

```{r}
head(female_egg_data)

```


### Loading and cleaning data

In our previous chapters we worked through the steps of importing a dataset, standardising column names, formatting column types and standardising date formats. 

We also carried out simple exploratory steps using packages such as `skimr`. 

When we understand the requirements of our data - we can choose to convert code that runs on specific datasets (object-oriented programming) into something more abstract - functional programming. 

```{r, eval =FALSE}
data <- read_csv("data/raw/female_egg_data.csv") |> 
  clean_names() |> 
  mutate(date = dmy(date))

skim(data)

print(data)

```

:::{.task}
::::{.task-header}
Your turn
::::
::::{.task-container}

Using the Rstudio drop down Code > Extract Function
- you can highlight code and start turning it into a named function. 

- this is relatively simple and may not give you exactly what you want (AI support can help here)


::::
:::



When we have a series of functions - these can be run without modification on new data:

```{r}
load_and_clean_data <- function(path, 
                                date_col = NULL,
                                delim = ",",
                                show_skim = TRUE) {
  # Read data using readr
  df <- readr::read_delim(path, delim = delim, show_col_types = FALSE) |>
    janitor::clean_names()
  
  # Convert date column if provided
  if (!is.null(date_col) && date_col %in% names(df)) {
    df <- df |>
      dplyr::mutate({{ date_col }} := lubridate::dmy(.data[[date_col]]))
  } else if (!is.null(date_col)) {
    warning(paste("Column", date_col, "not found in data — skipping date conversion"))
  }
  
  # Optionally skim summary
  if (show_skim) {
    cat("\n📊 Quick data summary:\n")
    print(skimr::skim(df))
  }
  
  return(df)
}

```


```{r, eval = FALSE}
load_and_clean_data("data/raw/female_egg_data.csv")
```

:::{.callout-note}

In theory this function could be applied to any dataset (as long as it was in tidy format)

:::


### Validating data



```{r}
library(assertr)
library(tidyverse)
library(rlang)

data_checks <- function(data,
                        metadata_path = NULL,
                        bounded_cols = NULL,
                        bounds = c(1, 30),
                        treatment_levels = c("A", "B"),
                        just_warn = TRUE) {
  handler <- if (just_warn) assertr::just_warn else assertr::error_stop

  # Read metadata if provided
  metadata <- if (!is.null(metadata_path)) {
    readxl::read_excel(metadata_path, sheet = "Data Dictionary")
  } else {
    NULL
  }

  # Basic column presence check
  if (!is.null(metadata)) {
    required_cols <- metadata |> pull(Name)
    data <- data |> verify(has_all_names(required_cols), error_fun = handler)

    # Type check helper
    check_types <- function(d, m) {
      out <- tibble(
        variable = m$Name,
        expected = m$Type,
        actual   = map_chr(m$Name, ~ if (.x %in% names(d)) class(d[[.x]])[1] else NA_character_),
        match    = tolower(expected) == tolower(actual)
      )
      print(out)
      invisible(out)
    }

    check_types(data, metadata)
  }

  # Non-missing
  if (ncol(data) > 0) {
    data <- data |> assert(not_na, !!!syms(names(data)), error_fun = handler)
  }

  # Treatment
  if ("treatment" %in% names(data)) {
    data <- data |> assert(in_set(treatment_levels), treatment, error_fun = handler)
  }

  # Bounded numeric
  if (!is.null(bounded_cols)) {
    data <- data |> assert(within_bounds(bounds[1], bounds[2]), !!!syms(bounded_cols), error_fun = handler)
  }

  # Eggs logic check
  if (all(c("eggs_hatched", "eggs_laid") %in% names(data))) {
    data <- data |> verify(eggs_hatched <= eggs_laid, error_fun = handler)
  }

  return(data)
}


```


:::{.task}
::::{.task-header}
Question
::::
::::{.task-container}
Can we think of any ways this code could *potentially* be abstracted/improved further?

::::
:::

```{r, eval = FALSE}
  egg_data <- 
  load_and_clean_data("data/raw/female_egg_data.csv") |> 
  data_checks(metadata_path = "data-dictionary/insect_egg_metadata.xlsx")

```
