## Function based data checks {#sec-validate-functional}


```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))
```


```{r, include = FALSE}
library(dplyr)
library(assertr)

set.seed(123)
n <- 20

# Step 1: create dataset
female_egg_data <- tibble(
  female_id = 1:n,
  treatment = rep(c("A","B"), times =10),
  age_days = sample(1:30, n, replace = TRUE),
  eggs_laid = rpois(n, lambda = 50),
  eggs_hatched = pmin(eggs_laid, rpois(n, lambda = 45))
)

# Step 2: Introduce some deliberate issues
female_egg_data$age_days[1] <- 0                     # invalid age
female_egg_data$eggs_laid[2] <- mean(female_egg_data$eggs_laid) + 10*sd(female_egg_data$eggs_laid)  # extreme eggs_laid
female_egg_data$eggs_hatched[3] <- female_egg_data$eggs_laid[3] + 5  # eggs_hatched > eggs_laid

female_egg_data


```

## Motivation

We have run through the principles of organised data dictionaries and data validation. Once we understand the consistent types of data(sets) we encounter it is possible to start building data validation **pipelines**.

- Pipelines are collections of custom data functions that can apply consistent data cleaning steps to different data that fits a particular layout

- Pipelines can help speed up the process of data validation, analysis and plotting. 

### Examples

- Parameterised reports - simple summary reports that can be generated from a dataset with a button push

- Parameterised data import and validation pipelines - our second example and presented here:


## The data

In the previous chapters we worked through the process of applying data validation checks to a dataset on female age, fertility and fecundity according to treatments

```{r}
head(female_egg_data)

```


### Loading and cleaning data

In our previous chapters we worked through the steps of importing a dataset, standardising column names, formatting column types and standardising date formats. 

We also carried out simple exploratory steps using packages such as `skimr`. 

When we understand the requirements of our data - we can choose to convert code that runs on specific datasets (object-oriented programming) into something more abstract - functional programming. 

When we have a series of functions - these can be run without modification on new data:

```{r}
load_and_clean_data <- function(path, 
                                date_col = NULL,
                                delim = ",",
                                show_skim = TRUE) {
  # Read data using readr
  df <- readr::read_delim(path, delim = delim, show_col_types = FALSE) |>
    janitor::clean_names()
  
  # Convert date column if provided
  if (!is.null(date_col) && date_col %in% names(df)) {
    df <- df |>
      dplyr::mutate({{ date_col }} := lubridate::dmy(.data[[date_col]]))
  } else if (!is.null(date_col)) {
    warning(paste("Column", date_col, "not found in data — skipping date conversion"))
  }
  
  # Optionally skim summary
  if (show_skim) {
    cat("\n📊 Quick data summary:\n")
    print(skimr::skim(df))
  }
  
  return(df)
}

```



```{r}
load_and_clean_data("data dictionary/female_egg_data.csv")
```

:::{.callout-note}

In theory this function could be applied to any dataset (as long as it was in tidy format)

:::


### Validating data


```{r}
library(assertr)
library(dplyr)
library(rlang)

data_checks <- function(data,
                        metadata_path = NULL,
                        numeric_cols = NULL,
                        character_cols = NULL,
                        unique_cols = NULL,
                        bounded_cols = NULL,
                        bounds = c(1, 30),
                        treatment_levels = c("A", "B"),
                        just_warn = TRUE) {
  
  # Define error handler
  handler <- if (just_warn) assertr::just_warn else assertr::error_stop
  
  metadata <- readxl::read_excel(metadata_path, sheet = "Data Dictionary")
    
  required_cols <- metadata |>
      pull(`Name`)
  
  # Start pipeline
  out <- data |>
    verify(has_all_names(required_cols), error_fun = handler)
  
  # Numeric checks
  if (!is.null(numeric_cols)) {
    out <- out |>
      assert(is.numeric, !!!syms(numeric_cols), error_fun = handler)
  }
  
  # Character checks
  if (!is.null(character_cols)) {
    out <- out |>
      assert(is.character, !!!syms(character_cols), error_fun = handler)
  }
  
  # Unique checks
  if (!is.null(unique_cols)) {
    out <- out |>
      assert(is_uniq, !!!syms(unique_cols), error_fun = handler)
  }
  
  # Non-missing checks
  all_cols <- names(data)
  
  if (!is.null(all_cols)) {
    out <- out |>
      assert(not_na, !!!syms(all_cols), error_fun = handler)
  }
  
  # Treatment in set
  if ("treatment" %in% names(data)) {
    pred_treatment <- assertr::in_set(treatment_levels)
    out <- out |>
      assert(pred_treatment, treatment, error_fun = handler)
  }
  
  # Bounded numeric checks
  if (!is.null(bounded_cols)) {
    out <- out |>
      assert(within_bounds(bounds[1], bounds[2]), !!!syms(bounded_cols),
             error_fun = handler)
  }
  
  # Eggs logic check if present
  if (all(c("eggs_hatched", "eggs_laid") %in% names(out))) {
    out <- out |>
      verify(eggs_hatched <= eggs_laid, error_fun = handler)
  }
  
  return(out)
}


```


:::{.task}
::::{.task-header}
Question
::::
::::{.task-container}
Can we think of any ways this code could *potentially* be abstracted/improved further?

`r hide("Solution")`

`r unhide()`

::::
:::

```{r}
  egg_data <- 
  load_and_clean_data("data dictionary/female_egg_data.csv") |> 
  data_checks(metadata_path = "data dictionary/insect_egg_metadata.xlsx")

```
