# Dates {#sec-dates}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("data/penguin_clean_names.RDS")
```

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://github.com/Philip-Leftwich/Oct-Intro-Analytics/raw/refs/heads/dev/data/penguin_clean_names.RDS"))

```

:::


Working with dates can be tricky, treating date as strictly numeric is problematic, it won't account for number of days in months or number of months in a year. 

Additionally there's a lot of different ways to write the same date:

* 13-10-2019

* 10-13-2019

* 13-10-19

* 13th Oct 2019

* 2019-10-13

This variability makes it difficult to tell our software how to read the information, luckily we can use the functions in the `lubridate` package. 


:::{.callout-warning}
If you get a warning that some dates could not be parsed, then you might find the date has been inconsistently entered into the dataset.

Pay attention to warning and error messages
:::

Depending on how we interpret the date ordering in a file, we can use `ymd()`, `ydm()`, `mdy()`, `dmy()` 

## Reformat

## Extract

## Excel date formats

```{r}
library(janitor)

excel_numeric_to_date(42370)
```

## Examples

* **Question** What is the appropriate function from the above to use on the `date_egg` variable in the penguins dataset?


`r longmcq(c("ymd()", "ydm()", "mdy()", answer="dmy()"))`


`r hide("Solution")`


```{r, eval = T, warning = F, message = F}

penguins_clean_names <- penguins_clean_names |>
  mutate(date_egg = lubridate::dmy(date_egg))

```

`r unhide()`


Here we use the `mutate` function from `dplyr` to create a *new variable* called `date_egg_proper` based on the output of converting the characters in `date_egg` to date format. The original variable is left intact, if we had specified the "new" variable was also called `date_egg` then it would have overwritten the original variable. 


Once we have established our date data, we are able to perform calculations or extract information. Such as the date range across which our data was collected.  

### Calculations with dates

```{r, eval = F}
penguins_clean_names |> 
  summarise(min_date=min(date_egg),
            max_date=max(date_egg))
```

We can also extract and make new columns from our date column - such as a simple column of the year when each observation was made:

```{r, eval = F}
penguins_clean_names <- penguins_clean_names |> 
  mutate(year = lubridate::year(date_egg))

```

### Filter dates

```{r, eval = FALSE}
# return records after 2008
plants |>
  filter(date_egg >= ymd("2008-01-01"))
```

