# Duplicates {#sec-duplicate}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_clean_names <- readRDS("data/penguin_clean_names.RDS")
```

It is very easy when inputting data to make mistakes, copy something in twice for example, or if someone did a lot of copy-pasting to assemble a spreadsheet (yikes!). We can check this pretty quickly

:::{.callout-important collapse = "true"}

## Penguin clean names dataset

```{r, eval = FALSE}
penguins_clean_names <- readRDS(url("https://github.com/Philip-Leftwich/Oct-Intro-Analytics/raw/refs/heads/dev/data/penguin_clean_names.RDS"))

```

:::


## Duplicated rows


::: {.panel-tabset}

## dplyr

```{r, eval = F}
# check for whole duplicate 
# rows in the data
penguins_clean_names |> 
  filter(duplicated(across(everything())))
  sum() 

```

```
[1] 0
```


## janitor

```{r}
library(janitor)

penguins_clean_names |> 
  get_dupes()

```

:::

Great! 

If I did have duplications I could investigate further and extract these exact rows: 

```{r, eval = F}

# Inspect duplicated rows
penguins_clean_names |> 
  filter(duplicated(across(everything())))

```

```
A tibble:0 × 17
0 rows | 1-8 of 17 columns
```

```{r, eval = F}

# Keep only unduplicated data with !
penguins_clean_names |> 
  filter(!duplicated(across(everything())))

```

```{r, eval = T, echo = F}

# Keep only unduplicated data
penguins_clean_names |> 
  filter(duplicated(across(everything()))) |> 
  head()

```



### Counting unique entries


```{r}
penguins_clean_names |> 
  summarise(
  n = n(),
  n_distinct(individual_id)
  )

```




