# Equivalence testing {#sec-equivalence}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(easystats)
source("R/booktem_setup.R")
source("R/my_setup.R")

```


## Motivation

Imagine we’re running an experiment comparing **different insect lines or treatments** — for example, testing whether two genetic strains, or two rearing conditions, lead to *different body masses or survival rates*.

### The standard question

Traditionally, we analyze such data using null-hypothesis significance testing (NHST) and design our studies with power analysis.

- The null hypothesis (H₀) says there’s no difference between treatments (e.g., mean body mass difference = 0).

- A significant p-value (p < 0.05) leads us to conclude that there is a difference.

- Power analysis tells us how many samples we need to detect a given difference (e.g., a 5% or 10% change in mass).

This framework is excellent when our goal is to find differences — for example, when we hope a new treatment improves performance relative to a control.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(ggplot2)

set.seed(123)

# Define parameters
x <- seq(-4, 8, length.out = 1000)
y_null <- dnorm(x, mean = 0, sd = 1)      # H0 distribution
y_alt  <- dnorm(x, mean = 3, sd = 1)      # H1 distribution (true effect)

# Observed effect (e.g., from an experiment)
obs <- 1.8

# Build data frame
df <- data.frame(x, y_null, y_alt)

gg_nhst <- ggplot(df, aes(x)) +
    # --- Null distribution ---
    geom_area(aes(y = y_null), fill = "grey80", color = "grey50", alpha = 0.7) +
    # Acceptance region (typical central 95% of H0)
    stat_function(fun = dnorm, args = list(mean = 0, sd = 1),
                  geom = "area", fill = "lightblue", alpha = 0.4,
                  xlim = c(-1.96, 1.96)) +
    # --- Alternative distribution ---
    geom_line(aes(y = y_alt), color = "darkorange3", linewidth = 1.2) +
    geom_area(aes(y = y_alt), fill = "orange", alpha = 0.2) +
    # Observed effect
    geom_vline(xintercept = obs, color = "red", linewidth = 1.1, linetype = "solid") +
    annotate("text", x = obs, y = 0.05, label = "Observed effect", color = "red", hjust = -0.1) +
    # Labels for H0 and H1
    annotate("text", x = 0, y = 0.42, label = "H[0]", size = 6, color = "grey30") +
    annotate("text", x = 3, y = 0.42, label = "H[1]", size = 6, color = "darkorange3") +
    annotate("segment", x = -1.96, xend = 1.96, y = 0.02, yend = 0.02,
             arrow = arrow(length = unit(0.15, "cm")), color = "steelblue") +
    annotate("segment", x = 1.96, xend = -1.96, y = 0.02, yend = 0.02,
             arrow = arrow(length = unit(0.15, "cm")), color = "steelblue") +
    labs(
        title = "Classic Null Hypothesis Testing (NHST)",
        subtitle = "Observed effect falls inside the null acceptance region\n→ fail to reject H₀, but effect is more consistent with H₁",
        x = "Test statistic / Effect size",
        y = "Density"
    ) +
    theme_minimal(base_size = 14) +
    theme(panel.grid.minor = element_blank())

gg_nhst

```

### Testing for no difference

In many biological and applied contexts, the question isn’t “Are they different?” but rather:

> “Are they effectively the same?”

For example:

- We’ve developed a new insect diet and want to ensure it performs as well as the standard one.

- We’re testing whether two breeding lines have equivalent fitness before merging them in a production colony.

- We want to confirm that a new pesticide formulation is no worse than the established product within a biologically acceptable margin.

In these cases, a *non-significant* result from a classical test (p > 0.05) does not prove equivalence 

> it merely means we don’t have enough evidence to detect a difference.

The true difference could still be large; we just might not have had enough power.

## Equivalence testing

Equivalence testing inverts the logic:

- We define a Region of Practical Equivalence (ROPE) — a range of differences so small they are biologically or practically negligible (e.g., ±5% in body mass).

- The null hypothesis becomes:

    𝐻0: The true difference lies outside the ROPE (i.e., not equivalent)


- The alternative hypothesis:

    𝐻1: The true difference lies entirely within the ROPE (i.e., equivalent)


If the confidence interval for our treatment difference lies completely inside this equivalence range, we can *accept practical equivalence*.

```{r, echo = FALSE, message = F, warning = F}
# Simulated estimate distribution (like a posterior or sampling dist)
x2 <- seq(-4, 4, length.out = 1000)
y2 <- dnorm(x2, mean = 0.4, sd = 0.8)

rope <- c(-1, 1)

df2 <- data.frame(x2, y2)

gg_rope <- ggplot(df2, aes(x2, y2)) +
  # ROPE region
  annotate("rect", xmin = rope[1], xmax = rope[2],
           ymin = -Inf, ymax = Inf, fill = "orange", alpha = 0.2) +
  geom_area(fill = "steelblue", alpha = 0.5, color = "steelblue") +
  geom_vline(xintercept = 0, color = "red", linetype = "dashed") +
  annotate("text", x = 0, y = 0.4, label = "0 (no effect)", color = "red", vjust = -1) +
  annotate("text", x = 0, y = 0.18, label = "ROPE\n(negligible range)", color = "orange3") +
  labs(
    title = "Equivalence Testing (ROPE principle)",
    subtitle = "Most of the effect distribution lies within the ROPE \n→ evidence for practical equivalence",
    x = "Effect size",
    y = "Density"
  ) +
  theme_minimal(base_size = 14) +
  theme(panel.grid.minor = element_blank())

gg_rope


```

### What we can do with this

Equivalence testing allows us to:

- Formally conclude that two treatments perform “the same” within a tolerance, rather than merely failing to find a difference.

- Incorporate biological relevance (what counts as a meaningful difference) directly into the statistical test.

- Quantify how much of the observed uncertainty (the CI) lies inside or outside that practical range — via measures like the **Second-Generation p-Value (SGPV)**.

- Design experiments for “sufficient evidence of sameness” (safeguard power or TOST power), not just “evidence of difference.”

## Example:

We want to know if a new set of diet conditions effects the adult body mass of a set of insects: 

```{r, eval = TRUE, echo = FALSE}
downloadthis::download_link(
  
  link = "https://raw.githubusercontent.com/Philip-Leftwich/Oct-Intro-Analytics/refs/heads/dev/files/body_mass.RDS",
  button_label = "Download body mass data as RDS",
  button_type = "success",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)
```


```{r, echo = FALSE}
body_mass_data <- readRDS("files/body_mass.RDS")

m_main <- lm(body_mass ~ treatment + sex, data = body_mass_data)
```

```{r, eval = FALSE}
body_mass_data <- readRDS("body_mass.RDS")

m_main <- lm(body_mass ~ treatment + sex, data = body_mass_data)
```

:::{.task}
::::{.task-header}
Confidence Intervals
::::
::::{.task-container}

- Fit 95% Confidence to your models

`r hide("Solution")`

```{r}

broom::tidy(m_main, conf.int = T)

#confint(m_main)

```

`r unhide()`

We can use our calculated confidence intervals *of the difference* to make a statement about the likely true difference in body mass for insects between treatment A and B.

::::
:::

## Equivalence testing

In equivalence testing, we’re not checking whether two means are different, but whether their difference is small enough to be considered practically the same — that is, within some limits (say between −Δ and +Δ).

To do this, we run two one-sided tests:

- Test if the difference is **greater** than the lower limit (−Δ).

- Test if the difference is **less** than the upper limit (+Δ).

Each of those tests is done at significance level α (for example, α = 0.05).
For both conditions to hold, we need **both** one-sided tests to be significant.

Now, when we express this as a single confidence interval, we must account for the fact that we used two one-sided tests — one for each tail. Because α is “spent” on both sides, the combined confidence level becomes (1 – 2α).

So if α = 0.05, we get a 90% confidence interval, not 95%.

### Set a standard

Let’s say a difference *smaller than 40 g* is acceptable.

This could be because we think its biologically negligible *or*

it could be that Diet B saves us a lot of money and we are willing to make the tradeoff within certain bounds

For this data we  define a ROPE (Region of Practical Equivalence) as between 0-40 g weight lost.


#### Output example:

```{r}
library(effectsize)
equivalence_test(m_main, range = c(0,-40))

```

#### Interpretation:

- The treatment effect (−21.8 g) lies completely within the ROPE of ±40 g → practically equivalent.

- The sex effect (−203 g) lies far outside the ROPE → meaningful biological difference.

- The **SGPV (Second-Generation p-value)** quantifies how much of the CI overlaps the ROPE (1 = fully inside, 0 = outside).

So:

- Even though the treatment effect was “statistically significant” (p = 0.04), the equivalence test says it’s practically negligible for our margins.

###  Visualize your equivalence results

You can use the see package to visualize the CI and ROPE region.

```{r}
library(see)
plot(equivalence_test(m_main, range = c(0,-40)))
```

This shows: 

- Each coefficient and its 90% CI

`r hide("90% CI")`

```{r}

broom::tidy(m_main, 
            conf.int = T, 
            conf.level = 0.9)

```

`r unhide()`

- The ROPE zone (here, ±40 g)

- A colour-coded decision (“Accepted”, “Rejected”, “Undecided”)

:::{.task}
::::{.task-header}
Try a different equivalence margin
::::
::::{.task-container}

- Are the diets equivalent at 0-20g?

`r hide("Solution")`

```{r}

equivalence_test(m_main, range = c(0,-20))
```

```{r}
plot(equivalence_test(m_main, range = c(0,-20)))

```

`r unhide()`

::::
:::

## Reporting

A linear model predicting insect body mass revealed a small treatment effect (−21.8 g, 90% CI [−39.4, −4.2]), and a strong effect of sex (−202.9 g, 90% CI [−220.5, −185.3]).

Equivalence testing using a ROPE of ±40g (A 10% difference in mass) indicated that the treatment difference was practically equivalent to zero, whereas the sex difference was meaningful.


## Glossary

| **Term**                                   | **Definition**                                                                                                                                                                                                                                  |
| ------------------------------------------ | ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- |
| **Null Hypothesis (H₀)**                   | The default assumption that there is *no true difference* between groups or treatments (e.g., mean body mass difference = 0).                                                                                                                   |
| **Alternative Hypothesis (H₁)**            | The competing claim that there *is* a true difference (e.g., H₁: difference ≠ 0).                                                                                                                                                               |
| **Region of Practical Equivalence (ROPE)** | A range of effect sizes (e.g., −5% to +5%) considered *biologically or practically negligible*. If the CI of an effect lies entirely within this region, the treatments are deemed practically equivalent.                                      |
| **Equivalence Testing**                    | A statistical approach that tests whether an observed effect is *small enough* to be considered negligible. Instead of asking “is there a difference?”, it asks “is any difference practically zero?”.                                          |
| **TOST (Two One-Sided Tests)**             | A frequentist method for equivalence testing that performs two one-sided hypothesis tests—one for each bound of the ROPE—to check whether the effect is significantly greater than the lower bound and smaller than the upper bound.            |
| **Second-Generation p-Value (SGPV)**       | A measure of how much of the confidence interval lies within the ROPE. Values: 1 → entirely inside (strong evidence for equivalence); 0 → entirely outside (strong evidence against equivalence); between 0–1 → partial overlap (inconclusive). |
| **Confidence Interval (CI)**               | A range that likely contains the true parameter value with a specified probability (e.g., 90% or 95%). In equivalence testing, the CI is compared directly to the ROPE.                                                                         |
| **Effect Size**                            | A quantitative measure of the magnitude of a treatment effect (e.g., *Cohen’s d*, *η²*, or regression coefficient). Equivalence testing uses effect size thresholds to define what is “small enough” to be negligible.                          |
| **Statistical Significance (p-value)**     | The probability of observing an effect as extreme as the current result if H₀ were true. A small p-value (< 0.05) implies evidence against H₀, but does *not* imply equivalence when non-significant.                                           |
| **Power (1 − β)**                          | The probability that a test correctly rejects a false null hypothesis — i.e., detects a true effect of a specified size. Classic power analysis is designed for detecting differences, not for confirming equivalence.                          |
| **Practical Equivalence**                  | The conclusion that two treatments perform similarly enough that any difference is negligible for practical or biological purposes.                                                                                                             |
| **Undecided Region**                       | When the CI overlaps both the ROPE and values outside it, indicating insufficient evidence to conclude either equivalence or difference.                                                                                                        |
