# Themes {#sec-ggplot3}

```{r, include = FALSE}
library(tidyverse)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")

penguins <- readRDS(here("files", "chapter5.RDS"))

#___________________________----
# SET UP ----
# An analysis of the bill dimensions of male and female 
# Adelie, Gentoo and Chinstrap penguins

# Data first published in  Gorman, KB, TD Williams, and WR Fraser. 
# 2014. 
# “Ecological Sexual Dimorphism and Environmental Variability 
# Within a Community of Antarctic Penguins (Genus Pygoscelis).” 
# PLos One 9 (3): e90081.
# https://doi.org/10.1371/journal.pone.0090081. 
#__________________________----

# PACKAGES ----
library(tidyverse) # tidy data packages
library(here) # organised file paths
library(janitor) # cleans variable names
#__________________________----

# IMPORT DATA ----
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))

# check the data has loaded, prints first 10 rows of dataframe
penguins_raw
#__________________________----
# CLEAN DATA ----

# clean all variable names to snake_case
# using the clean_names function from the janitor package
# note we are using assign <-
# to overwrite the old version of penguins
# with a version that has updated names
# this changes the data in our R workspace
# but NOT the original csv file

# clean the column names
# assign to new R object
penguins_clean <- janitor::clean_names(penguins_raw)

# quickly check the new variable names
colnames(penguins_clean)

# shorten the variable names for N and C isotope blood samples

penguins_clean <- rename(penguins_clean,
  "delta_15n" = "delta_15_n_o_oo", # use rename from the dplyr package
  "delta_13c" = "delta_13_c_o_oo"
)

# use mutate and case_when for a statement that conditionally changes the names of the values in a variable
penguins <- penguins_clean |>
  mutate(species = case_when(
    species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
    species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo",
    species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap"
  ))

# use mutate and if_else
# for a statement that conditionally changes
# the names of the values in a variable
penguins <- penguins |>
  mutate(sex = if_else(
    sex == "MALE", "Male", "Female"
  ))

# use lubridate to format date and extract the year
penguins <- penguins |>
  mutate(date_egg = lubridate::dmy(date_egg))

penguins <- penguins |>
  mutate(year = lubridate::year(date_egg))



```


## Learning Objectives

By the end of this workshop, you will be able to:

- Apply custom themes that fit with brand design choices

In professional settings, visual outputs should communicate not only data but identity.
This session shows how to translate a set of visual branding principles (colours, typography, tone) into a coherent `ggplot2` theme and apply it consistently across plot types.



## Making a custom palette

Establishing a coherent colour palette ensures every figure feels unified.

The first code block defines a named vector of brand colours, assigning meaningful names (teal, navy, yellow, etc.) to specific hex codes.

This allows the colours to be referenced consistently across plots:


```{r}

brand_colours <- c(
  "teal"    = "#09baba",  # bright teal
  "green"   = "#0d8242",  # dark green
  "lime"    = "#0ee565",  # vivid lime
  "navy"    = "#2c4977",  # deep navy
  "blue"    = "#3293BD",  # bright blue
  "yellow"  = "#d2db2e"   # lime-yellow accent
)

mono_palette <- function(base_col = "#2c4977", n = 5) {
  colorRampPalette(c(base_col, "white"))(n)
}

# Example use
mono_blue <- mono_palette( "#2c4977", n = 5)


brand_palettes <- list(
  brand_colours = brand_colours,
  discrete = c("#09baba", "#0d8242", "#3293BD", "#d2db2e", "#2c4977"),
  continuous_light = c("#2c4977","#09baba",  "#d2db2e" ),
  continuous_cool  = c("#09baba", "#3293BD", "#2c4977"),
  diverging        = c("#0d8242", "#d2db2e", "#3293BD"),
  mono_blue = mono_blue
)

show_brand_palettes <- function(palettes = brand_palettes) {
    df <- purrr::imap_dfr(palettes, ~{
        tibble(
            palette = .y,
            color = .x,
            order = seq_along(.x)
        )
    })
    
   plot <-  ggplot(df, aes(x = order, y = palette, fill = color)) +
        geom_tile(color = "white", linewidth = 0.8) +
        geom_text(aes(label = color), color = "black", size = 3, vjust = 2) +
        scale_fill_identity() +
        theme_minimal(base_size = 12) +
        theme(
            axis.title = element_blank(),
            axis.ticks = element_blank(),
            panel.grid = element_blank(),
            axis.text.x = element_blank(),
            axis.text.y = element_text(face = "bold")
        )
   plot
}

```


### Design check:

Which colours carry the brand’s main tone?

Which are suitable for accents or backgrounds?

Which combinations maintain legibility and contrast?

```{r}

show_brand_palettes()

```

`r hide("Colour Accessibility")`
```{r}

colorBlindness::cvdPlot(show_brand_palettes())

```

`r unhide()`


## Using showtext

In R, system fonts can vary across computers — meaning your carefully chosen typography might not appear the same for others or when exporting plots.
The `showtext` package solves this by rendering text using font files directly, ensuring consistent appearance across systems and outputs (e.g. PNG, PDF).

The simplest way to activate it is with `showtext::showtext_auto()` — once turned on, it applies automatically to all your plots. Check out (https://fonts.google.com/) for available fonts

```{r}
library(showtext)

# Register and activate a Google font
font_add_google("Monsieur La Doulaise", "Mon")

showtext_auto()

# Use the font in ggplot
ggplot(penguins, 
       aes(x = culmen_length_mm, 
           y = body_mass_g, 
           fill = species,
           shape = species))+
    geom_point()+
  ggtitle("A Fancy Scatterplot of Culmen length against Body Mass")+
  theme_bw(base_family ="Mon",
           base_size = 30)


showtext_auto(enable = FALSE) # to stop auto rendering
```

`showtext_auto()` makes font rendering automatic and portable — ensuring that your branded typeface appears consistently across systems, exports, and collaborators’ machines.

## Construct a base theme

```{r}

theme_brand <- function(base_size = 14, base_family = "Open Sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      # Text
      text = element_text(colour = mono_blue[1]),
      plot.title = element_text(face = "bold", size = rel(1.4), colour = mono_blue[1]),
      plot.subtitle = element_text(colour = mono_blue[2], size = rel(1.1)),
      plot.caption = element_text(size = rel(0.8), colour = mono_blue[2]),
      axis.title = element_text(face = "bold", colour = mono_blue[2]),
      axis.text = element_text(colour = mono_blue[3]),
      
      # Panels and gridlines
      panel.grid.major = element_line(colour = "grey90"),
      panel.grid.minor = element_blank(),
      panel.background = element_rect(fill = "white", colour = NA),
      
      # Facets
      strip.background = element_rect(fill = brand_colours["teal"], colour = NA),
      strip.text = element_text(face = "bold", colour = "white"),
      
      # Legends
      legend.title = element_text(face = "bold"),
      legend.position = "right"
    )
}


```


:::{.panel-tabset}

## Scatterplot

```{r}
ggplot(penguins, 
       aes(x = culmen_length_mm, 
           y = body_mass_g, 
           fill = species,
           shape = species))+
    geom_point()+
    labs(title= "Scatterplot", 
         subtitle = "Culmen Length (mm) against Body Mass (g)", 
         caption = "made by Philip Leftwich") +
  scale_shape_manual (values = c(21,22,23))

```

## Applied theme

```{r}
ggplot(penguins, 
       aes(x = culmen_length_mm, 
           y = body_mass_g, 
           fill = species,
           shape = species))+
    geom_point()+
    labs(title= "Scatterplot", 
         subtitle = "Culmen Length (mm) against Body Mass (g)", 
         caption = "made by Philip Leftwich") +
    theme_brand() +
  scale_shape_manual (values = c(21,22,23))

```

:::

- Which aspects make this “on-brand”?

- Is the hierarchy of colour, text, and space clear?

## Making an advanced custom theme

A theme function defines the non-data aspects of a plot: typography, line styles, gridlines, and overall layout.

Here - `theme_brand()` creates a cohesive institutional style aligned with the brand colours.

Key elements to note:

- Typography – the function attempts to load the Google font Open Sans via the `showtext` package @R-showtext, ensuring reproducible fonts across systems.

- Axis options – axis_type toggles between continuous and categorical grid settings, adjusting gridlines and spacing accordingly.

- Colour palettes – palette selects one of the brand palette variants and attaches appropriate scale_fill_* and scale_colour_* definitions automatically.

Because the function returns both a theme object and corresponding colour scales, you can call it directly within a ggplot pipeline and immediately inherit both the look and the colour logic:

```{r}

#' Create a Branded ggplot2 Theme with Integrated Colour Palettes
#'
#' @description
#' Builds a ggplot2 theme styled according to institutional or project-specific 
#' brand aesthetics. The function defines typographic, gridline, and layout styles 
#' while automatically applying brand colour scales for both discrete and continuous data. 
#' 
#' This function is designed for consistent, reproducible brand alignment across plots.
#'
#' @param base_size Numeric. Base font size for the theme text elements. Default is `30`.
#' @param base_family Character. Font family to use (default is `"Open Sans"`). 
#' If available, the function attempts to load it automatically via the `showtext` package.
#' @param base_line_size Numeric. Line width for base line elements. Default is `0.8`.
#' @param base_rect_size Numeric. Line width for rectangular elements (panel borders, etc.). Default is `0.5`.
#' @param axis_type Character. Determines gridline layout: `"categorical"` for discrete x-axes 
#' or `"continuous"` for numeric axes. Default is `"categorical"`.
#' @param palette Character. Selects which brand palette variant to apply. 
#' Options are `"discrete"`, `"continuous_light"`, `"continuous_cool"`, or `"diverging"`. 
#' Default is `"discrete"`.
#' @param midpoint Numeric. Midpoint value used for diverging colour scales. Default is `0`.
#'
#' @details
#' The function loads and activates the Google font *Open Sans* (if available) 
#' using `showtext` and `sysfonts`. It composes the theme from a base minimal layout, 
#' adjusting typography, gridlines, and strip backgrounds to reflect the brand palette.
#'
#' The function returns a list containing the ggplot2 theme object and 
#' corresponding colour/fill scales. When added to a ggplot pipeline, both are applied automatically:
#'
#' ```r
#' ggplot(data, aes(x, y, colour = group)) +
#'   geom_point() +
#'   theme_brand(axis_type = "continuous", palette = "discrete")
#' ```
#'
#' @return
#' A list containing:
#' \itemize{
#'   \item A ggplot2 theme object controlling non-data aesthetics.
#'   \item Corresponding `scale_*` definitions for colour and fill, matched to the chosen palette.
#' }
#'
#' @examples
#' \dontrun{
#' ggplot(penguins, aes(bill_length_mm, bill_depth_mm, colour = species)) +
#'   geom_point(size = 3) +
#'   labs(title = "Penguin bill dimensions") +
#'   theme_brand(axis_type = "continuous", palette = "discrete")
#' }
#'
#' @seealso [ggplot2::theme()], [ggplot2::scale_color_manual()], [showtext::showtext_auto()]
#'
#'
#' @dependencies
#' Requires: `ggplot2`, `showtext`, `sysfonts`, and user-defined objects 
#' `brand_colours`, `brand_palettes`, and `mono_blue` to be available in the environment.

# ---- Theme Function ----
theme_brand <- function(base_size = 30,
                             base_family = "Open Sans",
                             base_line_size = 0.8,
                             base_rect_size = 0.5,
                             axis_type = c("categorical", "continuous"),
                             palette = c("discrete", "continuous_light", "continuous_cool", "diverging"),
                             midpoint = 0) {

  palette <- match.arg(palette)
  
  axis_type <- match.arg(axis_type) 

  # Load and activate font via showtext
  if (requireNamespace("showtext", quietly = TRUE)) {
    showtext::showtext_auto(enable = TRUE)
    # Add Google font if not already registered
    if (!("Open Sans" %in% sysfonts::font_families())) {
      try(sysfonts::font_add_google("Open Sans", "Open Sans"), silent = TRUE)
    }
  }

   # ---- Core Theme ----
  base <- ggplot2::theme(
    line = ggplot2::element_line(colour = "#000000", linewidth = base_line_size),
    rect = ggplot2::element_rect(fill = "#FFFFFF", colour = "#000000", linewidth = base_rect_size),
    text = ggplot2::element_text(family = base_family, colour = "#000000", size = base_size),

    plot.title = ggplot2::element_text(face = "bold", hjust = 0, size = rel(1.8),margin = ggplot2::margin(12, 0, 8, 0),
                                       colour = mono_blue[1]),
    plot.subtitle = ggplot2::element_text(hjust = 0, size = base_size * 1.2,margin = ggplot2::margin(4, 0, 0, 0),
                                          colour = mono_blue[2]),
    plot.caption = ggplot2::element_text(hjust = 1, size = rel(0.9),margin = ggplot2::margin(8, 0, 0, 0),
                                         colour = mono_blue[3]),
    axis.text.y = element_text(colour = mono_blue[3], 
                               size = rel(0.8)),
    axis.title.y = element_text(colour = mono_blue[2],
                                margin = margin(0, 4, 0, 0)),
    axis.text.x = element_text(colour = mono_blue[3]
                               ),
    axis.title.x = element_text(colour = mono_blue[2],
                                margin = margin(0, 4, 0, 0)),
    panel.background = ggplot2::element_rect(fill = "white", colour = NA),
    strip.background = ggplot2::element_rect(fill = brand_colours["teal"], colour = NA),
    strip.text = ggplot2::element_text(face = "bold", colour = "white",
                                       size = rel(1.1)),
    panel.border = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(linewidth = base_line_size, colour = "black"),  
    axis.line.x = ggplot2::element_line(linewidth = base_line_size, linetype = "solid", colour = "black"),
    axis.line.y = ggplot2::element_line(linewidth = base_line_size, linetype = "solid", colour = "black")
  )
    
      # Axis-specific theme
  if (axis_type == "continuous") {
      axis_theme <- ggplot2::theme(
      panel.grid.major.y = ggplot2::element_line(colour = "#dedddd"),
      panel.grid.major.x = ggplot2::element_line(colour = "#dedddd"),
      legend.position = "right"
    )

  } else {
    axis_theme <- ggplot2::theme(
      panel.grid.major.y = ggplot2::element_line(colour = "#dedddd"),
      legend.position = "right"
    )
  }
    t <- list(base, axis_theme)
  

  # ---- Add Colour Scales ----
  s <- switch(palette,
    discrete = list(
      ggplot2::scale_color_manual(values = brand_palettes$discrete),
      ggplot2::scale_fill_manual(values = brand_palettes$discrete)
    ),
    continuous_light = list(
      ggplot2::scale_color_gradientn(colours = brand_palettes$continuous_light),
      ggplot2::scale_fill_gradientn(colours = brand_palettes$continuous_light)
    ),
    continuous_cool = list(
      ggplot2::scale_color_gradientn(colours = brand_palettes$continuous_cool),
      ggplot2::scale_fill_gradientn(colours = brand_palettes$continuous_cool)
    ),
    diverging = list(
      ggplot2::scale_color_gradient2(
        low = brand_palettes$diverging[1],
        mid = brand_palettes$diverging[2],
        high = brand_palettes$diverging[3],
        midpoint = midpoint
      ),
      ggplot2::scale_fill_gradient2(
        low = brand_palettes$diverging[1],
        mid = brand_palettes$diverging[2],
        high = brand_palettes$diverging[3],
        midpoint = midpoint
      )
    )
  )

  c(list(t), s)
}


```


:::{.panel-tabset}

## Scatter plot
```{r}

ggplot(penguins, 
       aes(x = culmen_length_mm, 
           y = body_mass_g, 
           fill = species,
           shape = species))+
    geom_point()+
    labs(title= "Scatterplot", 
         subtitle = "Culmen Length (mm) against Body Mass (g)", 
         caption = "made by Philip Leftwich") +
  scale_shape_manual (values = c(21,22,23))

```

## Theme

```{r}
ggplot(penguins, 
       aes(x = culmen_length_mm, 
           y = body_mass_g, 
           fill = species,
           shape = species))+
    geom_point()+
    labs(title= "Scatterplot", 
         subtitle = "Culmen Length (mm) against Body Mass (g)", 
         caption = "made by Philip Leftwich") +
    theme_brand(axis_type  = "continuous") +
  scale_shape_manual (values = c(21,22,23))

```

:::



:::{.panel-tabset}


## Boxplot

```{r}

dodge <- 0.8

penguins |> 
  drop_na(sex) |> 
ggplot(    aes(x = species, 
           y = body_mass_g, 
           fill = sex))+
  geom_violin(position = position_dodge(width = dodge), alpha = .4)+
  geom_boxplot(position = position_dodge(width = dodge), width = .5, outliers = FALSE, show.legend = FALSE)+
  labs(title= "Boxplot", 
       subtitle = "A comparison of Body Mass (g) by Sex and Species", 
       caption = "made by Philip Leftwich") 

```

## Theme

```{r}

dodge <- 0.8

penguins |> 
  drop_na(sex) |> 
ggplot(    aes(x = species, 
           y = body_mass_g, 
           fill = sex))+
  geom_violin(position = position_dodge(width = dodge), alpha = .4)+
  geom_boxplot(position = position_dodge(width = dodge), width = .5, outliers = FALSE, show.legend = FALSE)+
  labs(title= "Boxplot", 
       subtitle = "A comparison of Body Mass (g) by Sex and Species", 
       caption = "made by Philip Leftwich") +
    theme_brand(axis_type  = "categorical")

```

:::


:::{.task}
::::{.task-header}
Task
::::
::::{.task-container}

* [ ]  Create a `.R` file - save the funcitons here and give it a clear and recognisable filename

* This file when run now defines A named colour palette and a base theme function (theme_[yourbrand]())

* [ ] Try applying this theme to any plots you have/want to make.

::::
:::

Right now, your theme function is a standalone script — usable only within the current project.
To ensure it can be shared, versioned, and applied consistently across analyses and colleagues, the correct next step is to encapsulate it in an **R package**

