# ggplot {#sec-ggplot}

```{r, include = FALSE}
library(tidyverse)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
#source("R/clean_penguins.R")

#___________________________----
# SET UP ----
# An analysis of the bill dimensions of male and female 
# Adelie, Gentoo and Chinstrap penguins

# Data first published in  Gorman, KB, TD Williams, and WR Fraser. 
# 2014. 
# “Ecological Sexual Dimorphism and Environmental Variability 
# Within a Community of Antarctic Penguins (Genus Pygoscelis).” 
# PLos One 9 (3): e90081.
# https://doi.org/10.1371/journal.pone.0090081. 
#__________________________----

# PACKAGES ----
library(tidyverse) # tidy data packages
library(here) # organised file paths
library(janitor) # cleans variable names
#__________________________----

# IMPORT DATA ----
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))

# check the data has loaded, prints first 10 rows of dataframe
penguins_raw
#__________________________----
# CLEAN DATA ----

# clean all variable names to snake_case
# using the clean_names function from the janitor package
# note we are using assign <-
# to overwrite the old version of penguins
# with a version that has updated names
# this changes the data in our R workspace
# but NOT the original csv file

# clean the column names
# assign to new R object
penguins_clean <- janitor::clean_names(penguins_raw)

# quickly check the new variable names
colnames(penguins_clean)

# shorten the variable names for N and C isotope blood samples

penguins_clean <- rename(penguins_clean,
  "delta_15n" = "delta_15_n_o_oo", # use rename from the dplyr package
  "delta_13c" = "delta_13_c_o_oo"
)

# use mutate and case_when for a statement that conditionally changes the names of the values in a variable
penguins <- penguins_clean |>
  mutate(species = case_when(
    species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
    species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo",
    species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap"
  ))

# use mutate and if_else
# for a statement that conditionally changes
# the names of the values in a variable
penguins <- penguins |>
  mutate(sex = if_else(
    sex == "MALE", "Male", "Female"
  ))

# use lubridate to format date and extract the year
penguins <- penguins |>
  mutate(date_egg = lubridate::dmy(date_egg))

penguins <- penguins |>
  mutate(year = lubridate::year(date_egg))



```


## Learning Objectives

By the end of this chapter, you will be able to:

- Prepare and summarise data for advanced plotting.

- Control aesthetics and grouping behaviour in multilayered plots.

- Use and combine a broad range of geoms, including line, text, and label.

- Customise and extend colours, palettes, scales, axes, and legends.

- Use facets effectively for multi-dimensional comparisons.

- Arrange and enhance plots using patchwork and marginal plots.

- Export and enhance visualisations with camcorder and plotly.

## Grouping logic and plots

`ggplot2` automatically groups data based on discrete variables in your aesthetics (like colour or fill). 

```{r}
ggplot(penguins, aes(x = flipper_length_mm, y = body_mass_g, color = species)) +
geom_line(aes(group = species), linewidth = 1.2) +
theme_minimal()
```

- **Implicit grouping**: When colour or fill is mapped to a categorical variable.

- **Explicit grouping**: When you use group = inside aes().

```{r}
ggplot(penguins, aes(x = year, y = body_mass_g, color = species)) +
geom_line(linewidth = 1.2) +
scale_x_continuous(breaks = c(2007:2008), limits = c(2007,2008))+
theme_minimal()

```

```{r}


ggplot(penguins, aes(x = year, y = body_mass_g, color = species)) +
geom_line(aes(group = individual_id), linewidth = 1.2) +
scale_x_continuous(breaks = c(2007:2008), limits = c(2007,2008))+
theme_minimal()
```

## Exploring Common Geom Types

We’ve seen scatterplots, boxplots, and histograms—but ggplot offers many more geoms for storytelling.

| Data type              | Example geom                                  | Purpose              |
|------------------------|-----------------------------------------------|----------------------|
| Continuous–continuous  | `geom_point()`, `geom_line()`, `geom_smooth()` | Relationships/trends |
| Categorical–continuous | `geom_boxplot()`, `geom_violin()`, `geom_col()` | Comparisons          |
| Continuous             | `geom_histogram()`, `geom_density()`           | Distributions        |


### Combining geoms and layers

## Understanding Aesthetics

Aesthetics declared in ggplot(aes()) apply to all layers unless overridden. This is known as aesthetic inheritance.

```{r}
ggplot(penguins, aes(culmen_length_mm, culmen_depth_mm)) +
geom_point(aes(color = species)) +
geom_smooth(aes(color = species), se = FALSE)
```

```{r}
ggplot(penguins, aes(culmen_length_mm, culmen_depth_mm)) +
geom_point(aes(color = species)) +
geom_smooth(color = "black", se = FALSE)
```

## Colours, palettes and extensions

```{r}
ggplot(penguins, aes(x = body_mass_g, fill = species)) +
geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
scale_fill_manual(values = c("Adelie" = "#1b9e77", "Chinstrap" = "#d95f02", "Gentoo" = "#7570b3")) +
theme_minimal()
```


```{r}
library(colorspace)

ggplot(penguins, aes(x = body_mass_g, fill = species)) +
geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
scale_fill_discrete_qualitative(palette = "Dark2") +
theme_minimal()


```

Wes Anderson palettes

https://www.datanovia.com/en/blog/easy-way-to-expand-color-palettes-in-r/

https://www.cararthompson.com/talks/nhsr2022-palatable-palettes/index.html



## Scales, Axes and Ordering


## Legend Mastery

## Fonts

## Add labels to plots - 5027A

The labs() function allows you to add and customize various labels in your plot, including:

Title: A main heading that summarizes the plot’s purpose.

Axis Labels: Descriptive labels for the x- and y-axes.

Caption: An optional caption to provide additional context or credit.

Legend Title: A title for the legend if you’re using color, shape, or fill aesthetics.

## Facets

## Patchwork


## Marginal plots


## Plotly


## Exporting safely
