# ggplot {#sec-ggplot}

```{r, include = FALSE}
library(tidyverse)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
#source("R/clean_penguins.R")

#___________________________----
# SET UP ----
# An analysis of the bill dimensions of male and female 
# Adelie, Gentoo and Chinstrap penguins

# Data first published in  Gorman, KB, TD Williams, and WR Fraser. 
# 2014. 
# “Ecological Sexual Dimorphism and Environmental Variability 
# Within a Community of Antarctic Penguins (Genus Pygoscelis).” 
# PLos One 9 (3): e90081.
# https://doi.org/10.1371/journal.pone.0090081. 
#__________________________----

# PACKAGES ----
library(tidyverse) # tidy data packages
library(here) # organised file paths
library(janitor) # cleans variable names
#__________________________----

# IMPORT DATA ----
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))

# check the data has loaded, prints first 10 rows of dataframe
penguins_raw
#__________________________----
# CLEAN DATA ----

# clean all variable names to snake_case
# using the clean_names function from the janitor package
# note we are using assign <-
# to overwrite the old version of penguins
# with a version that has updated names
# this changes the data in our R workspace
# but NOT the original csv file

# clean the column names
# assign to new R object
penguins_clean <- janitor::clean_names(penguins_raw)

# quickly check the new variable names
colnames(penguins_clean)

# shorten the variable names for N and C isotope blood samples

penguins_clean <- rename(penguins_clean,
  "delta_15n" = "delta_15_n_o_oo", # use rename from the dplyr package
  "delta_13c" = "delta_13_c_o_oo"
)

# use mutate and case_when for a statement that conditionally changes the names of the values in a variable
penguins <- penguins_clean |>
  mutate(species = case_when(
    species == "Adelie Penguin (Pygoscelis adeliae)" ~ "Adelie",
    species == "Gentoo penguin (Pygoscelis papua)" ~ "Gentoo",
    species == "Chinstrap penguin (Pygoscelis antarctica)" ~ "Chinstrap"
  ))

# use mutate and if_else
# for a statement that conditionally changes
# the names of the values in a variable
penguins <- penguins |>
  mutate(sex = if_else(
    sex == "MALE", "Male", "Female"
  ))

# use lubridate to format date and extract the year
penguins <- penguins |>
  mutate(date_egg = lubridate::dmy(date_egg))

penguins <- penguins |>
  mutate(year = lubridate::year(date_egg))



```

## Learning Objectives

By the end of this chapter, you will be able to:

- Understand the basic structure of a ggplot and how it builds layer by layer.

- Map data variables to visual aesthetics like position, color, and shape.

- Add and customise plot elements such as labels, themes, and titles.

- Use facets to compare subgroups effectively.

- Apply best practices for clear, meaningful visualisations.


Data visualisation helps us see relationships, patterns, and surprises that tables of numbers can easily hide. In this chapter, we’ll explore how to use **ggplot2**{@R-ggplot2} — part of the tidyverse — to build clear, insightful graphics step by step.

We’ll use our dataset of penguin measurements to explore how different species vary in their culmen dimensions. You’ll learn how to go from a simple scatterplot to well-polished figures that communicate data stories effectively.

:::{.callout-tip}
If you are struggling to write the code for you ggplot, try loading the `esquisse` @R-esquisse add-in to produce a point-and-click interface that will help you build your plot and produce R code at the end. 

:::


## Your first ggplot

Let’s start with the basic structure of a ggplot:

```{r}
ggplot(data = penguins, aes(x = culmen_length_mm, y = culmen_depth_mm)) +
geom_point()

```

Each ggplot is built up from layers:

- `ggplot(data, aes(...))` defines the dataset and aesthetic mappings (which variables go on the x and y axes, color, size, etc.)

- `geom_point()` adds a layer of points to represent each observation.

## Adding Meaning with Color and Labels

We can add more information by mapping color to another variable, such as species:

```{r}
ggplot(penguins, aes(x = culmen_length_mm, y = culmen_depth_mm, color = species)) +
geom_point() +
labs(
title = "Bill dimensions by penguin species",
x = "Bill length (mm)",
y = "Bill depth (mm)",
color = "Species"
) +
theme_minimal()

```

Now we can instantly see three clusters, one for each species. The `labs()` function customizes labels and titles, and `theme_minimal()` applies a clean look.

:::{.callout-tip}

Add one change at a time and run your code — you’ll see how each layer contributes

:::


### Mini-Challenge

Add a subtitle to your plot using the `labs()` function. Try experimenting with different themes (e.g., `theme_classic()` or `theme_light()`).

## Controlling Aesthetics

Sometimes we want to control aspects of appearance manually. For example, we can change point size, shape, or transparency (`alpha`):

```{r}
ggplot(penguins, aes(x = culmen_length_mm, y = culmen_depth_mm, color = species)) +
geom_point(size = 3, alpha = 0.7) +
theme_minimal()

```

:::{.callout-tip}
Adjusting alpha helps with overplotting, where many points overlap. It can also help when we want de-emphasise parts of our plot. 
:::

### Mini challenge

Make a plot with `shape = sex` and `color = species`. How do these aesthetics help you see relationships between variables?

::: {.callout-warning}

Common pitfall: Don’t overuse color. Use it only when it adds clarity. :::

:::

## Comparing groups with Facets

Faceting creates small multiples — one plot per subgroup — using `facet_wrap()` or `facet_grid()`.

```{r}
ggplot(penguins, aes(x = culmen_length_mm, y = culmen_depth_mm, color = species)) +
geom_point() +
facet_wrap(~ species) +
theme_minimal()

```

### Minichallenge

Try faceting by sex instead of species. What do you notice?

## Going further

Once you're comfortable with scatterplots you can explore other *geoms*

- `geom_histogram()` for distributions

- `geom_density()` for density distributions

- `geom_boxplot()` for comparisons

- `geom_smooth()` for trend lines

For example:

```{r}
ggplot(penguins, aes(x = culmen_length_mm, fill = species)) +
geom_histogram(bins = 20, alpha = 0.7, position = "identity") +
theme_minimal()

```

:::{.callout-tip}
# Bins

A histogram is a type of bar plot used to display the distribution of a continuous variable by grouping data into intervals, or "bins," along the x-axis. Each bin represents a range of values, and the height of the bar shows how many observations fall within that range, providing a visual summary of the frequency distribution of the data.
:::

### Mini challenge

Try changing the number of bins for the histogram, how does it change?


## Summary

In this section, you learned how to:

- Build a ggplot layer by layer

- Map variables to aesthetics

- Customise plots with labels, themes, and colour

- Use facets for subgroup comparisons






