# Power analysis {#sec-power}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(easystats)
source("R/booktem_setup.R")
source("R/my_setup.R")

```

## Power

if you want to plan a study with 80% power, a power analysis will answer the question:

> “If I were to repeat my experiment many thousands of times, what sample size will allow me to reject the null hypothesis on 80% of these occasions?”


### Why do we need to understand statistical power?

When we compare treatments, statistical power determines how likely a study is to detect a real difference if one exists. Studies with low power risk producing results that are misleading or inconclusive. In practice, insufficient power exposes researchers to several key mistakes:

- **False negatives (Type II errors)**: real treatment effects go undetected, leading to the false conclusion that treatments are equivalent.

- **Higher false discovery rates**: significant results in underpowered studies are more likely to be false positives.

- **Unstable or inflated effect sizes**: The less power we have the more exaggerated the apparent effects of false discoveries. All estimates are unreliable.

- **Wasted resources**: underpowered research consumes effort and participants without yielding clear insight.

## Simulate! 

Once we committed to an equation with concrete parameter values, we can simulate data for a sample of, say `n = 100` samples

Simply write down the equation and add random normal noise with the `rnorm()` function. 

:::{.callout-note}
Note that the `rnorm` function takes the standard deviation (not the variance).
:::

We set a `seed` so that the generated random numbers are reproducible - and then make a dataset according to our understanding of the experiment:

```{r}
set.seed(38596)

# Sample size and factors
n <- 100
treatment <- factor(rep(c("A", "B"), each = n/2))      # 50 A, 50 B
sex       <- factor(rep(c("Male","Female"), times = n/2))

# --- Specify the true model on the MEAN scale ---
# Using default treatment (dummy) coding with baseline = treatment A, sex Male
beta_0      <- 400   # mean for (A, Male)
beta_trtB   <- -20    # difference (B - A) among Males when no interaction in the true DGP
beta_male <-  -200    # difference (Female - Male) in A when no interaction
beta_int    <-  0    # interaction: extra effect for (B,Female) beyond main effects

# Build the model matrix for ~ treatment * sex (includes intercept, both mains, and interaction)
X <- model.matrix(~ treatment + sex)

# Pack coefficients in the order of colnames(X)
coef_true <- c(beta_0, beta_trtB, beta_male)
names(coef_true) <- colnames(X)

# Mean for each observation implied by predictors
mu <- as.vector(X %*% coef_true)

# --- Residual SD goes here (this is where you include sd) ---
sigma <- 50
body_mass <- rnorm(n, mean = mu, sd = sigma)

dat <- data.frame(body_mass, treatment, sex)

```


:::{.callout-important}

Our simulated data must include:

- All expected values/ranges of our predictors

- An estimate of the mean effect of each predictor on our dependent variable

- An estimate of variance (noise) in our data

:::


### Variance

For sensible measures of variance we really *need* some ground truthing data - pilot studies or measures from non-treatment populations


### Check the data

```{r, include = FALSE}

set.seed(38596)

# Sample size and factors
n <- 100
treatment <- factor(rep(c("A", "B"), each = n/2))      # 50 A, 50 B
sex       <- factor(rep(c("Male","Female"), times = n/2))

# --- Specify the true model on the MEAN scale ---
# Using default treatment (dummy) coding with baseline = treatment A, sex Male
beta_0      <- 400   # mean for (A, Male)
beta_trtB   <- -20    # difference (B - A) among Males when no interaction in the true DGP
beta_male <-  -200    # difference (Female - Male) in A when no interaction
beta_int    <-  0    # interaction: extra effect for (B,Female) beyond main effects

# Build the model matrix for ~ treatment * sex (includes intercept, both mains, and interaction)
X <- model.matrix(~ treatment + sex)

# Pack coefficients in the order of colnames(X)
coef_true <- c(beta_0, beta_trtB, beta_male)
names(coef_true) <- colnames(X)

# Mean for each observation implied by predictors
mu <- as.vector(X %*% coef_true)

# --- Residual SD goes here (this is where you include sd) ---
sigma <- 50
body_mass <- rnorm(n, mean = mu, sd = sigma)

dat <- data.frame(body_mass, treatment, sex)

# Fit models
m_main <- lm(body_mass ~ treatment + sex, data = dat)   # main effects only

```

```{r, include = FALSE}

# saveRDS(dat, file = "body_mass.RDS")

```


Let’s plot our simulated data:

```{r}
ggplot(dat, aes(x=treatment, 
               y=body_mass, 
               fill = sex)) + 
  geom_boxplot(width = .1, 
               outlier.shape = NA,
               show.legend = FALSE,
               position = position_dodge(width = .2)) +
  geom_point( shape =21,
              position= position_jitterdodge(dodge.width =.2))

```


Let’s assume that we collected and analyzed this specific sample from our population – what would have been the results?

```{r}

# Fit models
m_main <- lm(body_mass ~ treatment + sex, data = dat)   # main effects only

summary(m_main)
```

As you can see, in this simulated sample (based on a specific seed), the treatment effect is marginally significant (p < .05). But how likely are we to detect an effect with a sample of this size?


## Doing the power analysis

Now we need to repeatedly draw many samples and see how many of the analyses would have detected the existing effect. To do this, we put the code from above into a function called `sim1`. 

We coded the function to return the focal p-value (as default):

- `n` defines the required sample size

- `treatment_effect` defines the treatment effect in the raw scale (i.e., reduction in body mass)

This is an example of "function building" - we have taken the code expressed early and repackaged it so we can make new datasets (from the same parameters) quickly: 


`r hide("Simulation function")`

```{r}

sim1 <- function(n=100, treatment_effect=-20, print=FALSE) {
# Sample size and factors
n <- n
treatment <- factor(rep(c("A", "B"), each = n/2))      # 50 A, 50 B
sex       <- factor(rep(c("Male","Female"), times = n/2))

# --- Specify the true model on the MEAN scale ---
# Using default treatment (dummy) coding with baseline = treatment A, sex Male
beta_0      <- 400   
beta_trtB   <- treatment_effect    # argument of interest
beta_male <-  -200    
beta_int    <-  0    

# Build the model matrix for ~ treatment * sex (includes intercept, both mains, and interaction)
X <- model.matrix(~ treatment + sex)

# Pack coefficients in the order of colnames(X)
coef_true <- c(beta_0, beta_trtB, beta_male)
names(coef_true) <- colnames(X)

# Mean for each observation implied by predictors
mu <- as.vector(X %*% coef_true)

# --- Residual SD goes here (this is where you include sd) ---
sigma <- 50
body_mass <- rnorm(n, mean = mu, sd = sigma)

dat <- data.frame(body_mass, treatment, sex)

# Fit models
m_main <- lm(body_mass ~ treatment + sex, data = dat)   # main effects only
  
  # this lm() call should be exactly the function that you use
  # to analyse your real data set
  p_value <- summary(m_main)$coefficients["treatmentB", "Pr(>|t|)"]
  
  # Can print all models if desired
  if (print==TRUE) print(summary(res))
  else return(p_value)
}

```

`r unhide()`


- We then use the `replicate` function to repeatedly call the `sim` function for 1000 iterations.

:::{.callout-note collapse = "true"}

## Why 1000?

We are carrying out Monte Carlo Simulations (repeated random sampling to predict the probability of different outcomes in uncertain systems)

Because each draw will be slightly different, to get a robust power analysis we need to do enough simulations that it stabilises.

The number required may be different for different simulations (smaller variance more consistent results)

So it might be that you can start with a lower number (100) while checking your simulation works and increase until it stabilises

:::


```{r}
set.seed(38596)

iterations <- 1000 # the number of Monte Carlo repetitions
n <- 100 # the size of our simulated sample

# now run the sim() function a 1000 times and store the p-values in a vector:
p_values <- replicate(iterations, sim1(n=100))
```

How many of our 1000 virtual samples would have found the effect?

```{r, message = FALSE}
tibble(p_values) |> ggplot(aes(x = p_values))+geom_histogram()+geom_vline(xintercept = 0.05, linetype = "dashed", colour ="red")

```

```{r}
table(p_values < .05)
```

Only **52%** of samples with the same size of n=100 result in a significant p-value.

## Sample size planning

Now we know that a sample size of 100 does not lead to a sufficient power. But what sample size would we need to achieve a power of at least 80%? (A standard threshold). In the simulation approach you need to test different n until you find the necessary sample size. We do this by wrapping the simulation code into a loop that continuously increases the n. We then store the computed power for each n.

```{r}

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(100, 300, by=20) # test ns between 100 and 300

```

In this simulation-based power analysis, we estimate the probability of detecting a true treatment effect **across different sample sizes**. Because the data are simulated, we can specify the “true” effect size and variance directly, allowing us to obtain clean, interpretable power estimates.

However, in real data analysis, this is rarely the case. The parameters that drive the simulation—particularly the effect size and residual variance—would have to be estimated from preliminary or pilot data. Such estimates are **inherently fragile**: they depend on the specific sample, measurement precision, and underlying model assumptions.

Compare our two estimates of power below: 

:::{.panel-tabset}

## Fragile power

```{r}

set.seed(38596)

result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect=-21.82))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})

ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()
```

Hence, with n = 160, (80 in each group), we have an 80% chance to detect the effect.

## Actual power

```{r}

result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})

ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()
  

```

**We don't actually have sufficient power until n = 200, (100 in each group), we have an 80% chance to detect the effect**

:::

:::{.callout-warning}

Power analyses based on estimates made from initial experiments are known as *post hoc power calculations* and are misleading and simply not informative for data interpretation. 

:::


## Safeguard power analysis

Our estimate of the population effect size is an educated guess based on the best information available to us (much like the written hypotheses we have for a study). It’s accuracy can have critical implications for the sample size required to attain a certain level of power. For this reason, researchers should routinely consider varying their effect size estimate and assessing the impact this has on their power analysis. If we are using sample effect sizes reported in the literature to estimate the population effect size, one excellent recommendation is to formally consider the uncertainty of sample effect sizes. That is, take into account the 95% CI on these reported effects.

([Perugini et al., 2014](https://doi.org/10.1177/1745691614528519)) suggest we take into account the uncertainty of the sample effect size. To do so, they suggest for the researcher to **compute the lower absolute limit of a 60% CI on the sample effect**. This value would then be used as a safeguard effect size and inputted into a power analysis. Why 60%? Well, the lower limit of a 60% CI corresponds to a one-sided 80% CI. As Perugini, Gallucci, and Costantini (2014) describe, this implies a 20% risk that that the true population effect size is lower than this limit, and correspondingly, 80% assurance that the population effect size is equal to or greater than this limit. We focus on the lower boundary because of the asymmetry of overestimating compared to underestimating the population effect size in a power analysis.

We will apply a **safeguard power analysis** that aims for the lower end of a two-sided 60% CI around the parameter of the treatment effect (the intercept is irrelevant). (Of course you can use any other value than 60%, but this is the value (tentatively) mentioned by the inventors of the safeguard power analysis.)

```{r}
beta_hat <- -21.819      # estimate
se <- 10.607            # standard error from model

z60 <- qnorm(1 - (1 - 0.60)/2)

ci_60_lower <- beta_hat - z60 * se
ci_60_upper <- beta_hat + z60 * se

c(ci_60_lower, ci_60_upper)

```

The *conservative* treatment difference to plan for is `r glue::glue({round(ci_60_upper,2)})`

```{r}
set.seed(38596)

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(100, 500, by=20) # test ns between 100 and 300


result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -12.89))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})
  
ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()

```

With that more conservative effect size assumption, we would need around **450 individuals, i.e., 225 per group**.


## Smallest Effect of Interest

Many methodologists argue that we should not power for the expected effect size, but rather for the smallest effect size of interest (**SESOI**). In this case, a non-significant result can be interpreted as “We accept the H0, and even if a real effect existed, it most likely is too small to be relevant”.

**We have to decide for ourselves what we value as the smallest effect size of interest:**


```{r}
set.seed(38596)

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(40, 300, by=20) # 


result_1 <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -40)) # 40/400
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})

result_2 <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -20)) #20/400
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})

result_3 <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -10)) # 10/100
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})

result <- bind_rows(
  result_1 %>% mutate(effect_size = "10% difference"),
  result_2 %>% mutate(effect_size = "5% difference"),
  result_3 %>% mutate(effect_size = "2.5% difference")
) |> mutate(effect_size = factor(effect_size,
                                 levels =c("2.5% difference", 
                                           "5% difference", 
                                           "10% difference") 
                                                    ))

# Plot
ggplot(result, aes(x = n, y = power)) +
  geom_point() +
  geom_line() +
  labs(x = "Sample size (n)",
       y = "Power")+
  geom_hline(yintercept = 0.8,
             linetype = "dashed",
             colour = "red")+
  facet_wrap(~effect_size)+
  scale_x_continuous(breaks = seq(40,280, by =40))

```

If we decide that we are happy ignoring any difference that has < 10% effect on body mass across treatments, then we can produce a reasonably powered experiment with only 50 individuals (25 per treatment). 

### A checklist for calculating power

1. Define the statistical model that you will use to analyze your data (in our case: a linear regression). This sometimes is called the data generating mechanism or the data generating model.

2. Find empirical information about the parameters in the model. These are typically mean values, variances, group differences, regression coefficients, or correlations (in our case: a mean value (before treatment), a group difference, and an error variance). These numbers define the assumed state of the population.

3. Program your regression equation, which consists of a systematic (i.e., deterministic) part and (one or more) random error terms.

4. Do the simulation:

    a. Repeatedly sample data from your equation (only the random error changes in each run, the deterministic part stays the same).

    b. Compute the index of interest (typically: the p-value of a focal effect).

    c. Count how many samples would have detected the effect (i.e., the statistical power).

    d. Tune your simulation parameters until the desired level of power is achieved.

### Increasing power

Typically we would aim to change experimental sample numbers until we reach desired power. 

But we could also imagine making changes to measurement accuracy, or perhaps insect rearing parameters that will reduce variance?

## Reading

[Post hoc power is not informative](https://pmc.ncbi.nlm.nih.gov/articles/PMC9452450/)

[Safeguard Power Analysis](https://journals.sagepub.com/doi/10.1177/1745691614528519)
