# Power analysis{#sec-power}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(easystats)
source("R/booktem_setup.R")
source("R/my_setup.R")

```


## Motivation

## Setup

```{r, eval = FALSE}
library(simr)

library(tidyverse)
library(lme4)
```


## A simple linear model

Detect a mean difference delta on the original scale. We vary per-group sample size.

### Explanation of approach:

- For each n (per group) we simulate nsim datasets with n observations per group.

- Data generating model: y ~ N(mu_control, sd) and y ~ N(mu_control + delta, sd) for treatment.

- Fit lm(y ~ group), extract p-value for the treatment coefficient, compute power.

```{r}
# Parameters
delta <- 5         # raw mean difference you want to detect
sd_pop <- 10       # residual SD
alpha <- 0.05
nsim <- 200        # simulations per sample size (increase for smoother curve)
n_grid <- seq(10, 200, by = 10)  # sample sizes per group

# Function: simulate power at one sample size
power_lm_at_n <- function(n, nsim, delta, sd_pop, alpha) {
  pvals <- replicate(nsim, {
    dat <- tibble(
      group = rep(c("control", "treat"), each = n),
      y = c(rnorm(n, mean = 50, sd = sd_pop),
            rnorm(n, mean = 50 + delta, sd = sd_pop))
    )
    coef_p <- summary(lm(y ~ group, data = dat))$coefficients["grouptreat","Pr(>|t|)"]
    coef_p
  })
  mean(pvals < alpha, na.rm = TRUE)
}

# Compute power curve using purrr
power_lm_df <- tibble(n = n_grid)  |> 
  mutate(power = map_dbl(n, ~ power_lm_at_n(.x, nsim, delta, sd_pop, alpha)))

# Plot
ggplot(power_lm_df, aes(x = n, y = power)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  labs(title = "Power curve: simple two-group linear model",
       subtitle = paste0("Detect raw mean difference = ", delta, ", SD = ", sd_pop),
       x = "Sample size per group", y = "Estimated power") +
  theme_minimal()

```

## A multicovariate model

Detect treatment effect while adjusting for covariates.

### Explanation:

- Data generating model: y = β0 + β_trt*trt + β_sex*sex + β_age*age + ε.

- We simulate sex (binary) and age (continuous). We fit lm(y ~ trt + sex + age) and test the trt coefficient.

- Use same simulation pattern as (1).

```{r}
# Parameters
delta <- 5            # treatment effect to detect on original scale
beta_sex <- 2         # effect of sex (male vs female)
beta_age <- 0.3       # effect per year of age
sd_error <- 10
nsim <- 200
n_grid <- seq(20, 200, by = 20)   # per-group n

power_lm_cov_at_n <- function(n_per_group, nsim, delta, beta_sex, beta_age, sd_error, alpha) {
  pvals <- replicate(nsim, {
    N <- 2 * n_per_group
    trt <- rep(c(0,1), each = n_per_group)
    sex <- rbinom(N, 1, 0.5)                  # 0/1
    age <- rnorm(N, mean = 40, sd = 12)      # continuous covariate
    mu <- 50 + delta*trt + beta_sex*sex + beta_age*age
    y <- rnorm(N, mean = mu, sd = sd_error)
    dat <- tibble(y = y, trt = factor(trt), sex = factor(sex), age = age)
    pval <- summary(lm(y ~ trt + sex + age, data = dat))$coefficients["trt1","Pr(>|t|)"]
    pval
  })
  mean(pvals < alpha, na.rm = TRUE)
}

power_lm_cov_df <- tibble(n = n_grid) |> 
  mutate(power = map_dbl(n, ~ power_lm_cov_at_n(.x, nsim, delta, beta_sex, beta_age, sd_error, alpha)))

ggplot(power_lm_cov_df, aes(x = n, y = power)) +
  geom_line() + geom_point() +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  labs(title = "Power curve: linear model with covariates",
       subtitle = "Model: y ~ trt + sex + age; detect treatment effect on raw scale",
       x = "Sample size per group", y = "Estimated power") +
  theme_minimal()


```

:::{.callout-note}

Including covariates that explain outcome variance usually increases power compared to an unadjusted analysis (if covariates are correlated with outcome).

You must simulate covariates realistically (prevalence, distribution) because they affect power.

:::

## A binomial model

Detect a change in probability from p0 to p1 (or set via OR).

### Explanation:

- For each n per group simulate binary outcomes rbinom(n,1,p0) and rbinom(n,1,p1).

- Fit glm(y ~ group, family = binomial), extract p-value (Wald z for group).

- Compute power as fraction p < α.

```{r}
# Parameters
p0 <- 0.30
p1_values <- seq(0.35, 0.7, by = 0.05)  # sweep treatment probabilities to produce curves
nsim <- 200
n_grid <- seq(20, 200, by = 20)
alpha <- 0.05

# Function: power for one (n, p1)
power_glm_at_n_p1 <- function(n, p0, p1, nsim, alpha) {
  pvals <- replicate(nsim, {
    y0 <- rbinom(n, 1, p0)
    y1 <- rbinom(n, 1, p1)
    dat <- tibble(y = c(y0, y1), group = factor(rep(c(0,1), each = n)))
    fit <- glm(y ~ group, data = dat, family = binomial)
    summary(fit)$coefficients["group1","Pr(>|z|)"]
  })
  mean(pvals < alpha, na.rm = TRUE)
}

# Build data frame of power values for plotting (grid over n and p1)
power_grid <- expand_grid(n = n_grid, p1 = p1_values) |> 
  mutate(power = map2_dbl(n, p1, ~ power_glm_at_n_p1(.x, p0, .y, nsim, alpha)))

# Plot one facet per p1
ggplot(power_grid, aes(x = n, y = power, color = as.factor(p1))) +
  geom_line() + geom_point() +
  facet_wrap(~ p1) +
  labs(title = "Power curves: logistic regression (binary outcome)",
       subtitle = paste0("Baseline p0 = ", p0, "; lines show p1 values"),
       x = "Sample size per group", y = "Estimated power") +
  theme_minimal()


```

:::{.callout-note}

- `pwr::power.prop.test()` gives an analytic approximation for two-proportion tests — good as cross-check. Simulating the exact `glm()` is best if you plan to use a logistic regression in analysis.

- Use larger `nsim` (1000–5000) for higher-grade precision.

:::

## GLMM

Power in clustered data (e.g., replicates, subjects) where observations are nested; detect fixed treatment effect while allowing between-cluster variance.

### Explanation / steps:

- Make a realistic template dataset that reflects cluster structure (clusters, observations per cluster).

- Fit a `glmer` (for binary) or `lmer` (for continuous) model to that template.

- Set the fixed effect (if needed) to the raw effect you want to detect (for example, by adjusting the fitted model coefficients via `fixef()` — but easier is to simulate the template so it already contains the desired effect).

- Use `simr::powerSim()` on the fitted model to estimate power. To explore larger numbers of clusters use `extend(..., along = "cluster", n = desired_clusters)` and then `powerSim()` again.

```{r}
# --- Define simulation parameters ---
n_clusters <- 10              # number of clusters (random intercepts)
n_per_cluster <- 2            # 2 rows per cluster: control & treatment
n_trials <- 20                # number of trials per treatment per cluster

# Define baseline probabilities and treatment effect
p_control <- 0.9              # baseline success probability
p_treat   <- 0.8              # treated success probability
sd_cluster <- .6               # between-cluster logit SD

# --- Compute on logit scale ---
logit <- function(p) log(p / (1 - p))
inv_logit <- function(x) plogis(x)

beta0 <- logit(p_control)     # intercept for control
beta_treat <- logit(p_treat) - beta0  # treatment effect on logit scale

# --- Build the data frame ---
your_data <- tibble(
  cluster = factor(rep(1:n_clusters, each = n_per_cluster)),
  treat   = rep(c(0, 1), times = n_clusters)
)

# --- Add random effects and simulate binomial outcomes ---
cluster_effects <- rnorm(n_clusters, mean = 0, sd = sd_cluster)

your_data <- your_data |> 
  mutate(
    eta = beta0 + beta_treat * treat + cluster_effects[as.numeric(cluster)],
    prob = inv_logit(eta),
    successes = rbinom(n(), size = n_trials, prob = prob),
    failures = n_trials - successes
  )




```

```{r}
# Fit a random-intercept logistic regression
# Fit a mixed-effects logistic regression on grouped data
glmm_model <- glmer(
  cbind(successes, failures) ~ treat + (1 | cluster),
  data = your_data,
  family = binomial
)

summary(glmm_model)


```


```{r}

# Estimate power for current design (10 clusters)
ps_current <- powerSim(glmm_model, test = fixed("treat", "z"), nsim = 20) # 200
summary(ps_current)


```


```{r}
# Extend along the cluster dimension to simulate more sites
glmm_extended <- extend(glmm_model, along = "cluster", n = 30)

# Generate a power curve varying number of clusters
p_curve_treat <- powerCurve(
  fit = glmm_extended,
  test = fixed("treat"),   # test for the fixed treatment effect
  along = "cluster",            # vary number of clusters
  nsim = 50                    # number of simulation replicates per size 100!!!
)


```


```{r}
# Visualize the power curve
plot(p_curve_treat) +
  ggtitle("Power curve for treatment effect (GLMM)") +
  xlab("Number of clusters") +
  ylab("Estimated Power (p < 0.05)") +
  theme_minimal(base_size = 13)


```



