# Power analysis {#sec-power}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
library(easystats)
source("R/booktem_setup.R")
source("R/my_setup.R")

```

## Effect size

Researchers often have been trained to think in standardized effect sizes, such as Cohen’s 


- d, 

-a correlation r

- or other indices such as f^2 or eta^2


In the simulation approach, we typical work on the raw scale of variables.

The raw effect size is simply the treatment effect on the original BDI scale (i.e., the group difference in the outcome variable). In our case we assume that the treatment lowers the BDI score by 6 points, on average. Defining the raw effect requires some domain knowledge – you need to know your measurement scale, and you need to know what the values (and differences between values) mean. In our example, a reduction of 6 BDI points means that the average patient moves from a moderate depression (23 points) to a mild depression (17 points). Working with raw effect sizes forces you to think about your actual data (instead of plugging in content-free default standardized effect sizes) and enables you to do plausibility checks on your simulation.

The standardized effect size relates the raw effect size to the unexplained error variance. In the two-group example, this can be expressed as Cohen’s d, which is the mean difference divided by the standard deviation (SD):


The standardized effect size always relates two components: The raw effect size (here: 6 points difference) and the error variance. Hence, you can increase the standardized effect size by (a) increasing the raw treatment effect, or (b) reducing the error variance.

$$
d = \frac{M_{\text{treat}} - M_{\text{control}}}{SD} 
= \frac{17 - 23}{117} 
= -0.55
$$


## Simulate! 

Once we committed to an equation with concrete parameter values, we can simulate data for a sample of, say `n = 100` samples


Simply write down the equation and add random normal noise with the `rnorm()` function. 

:::{.callout-note}
Note that the `rnorm` function takes the standard deviation (not the variance).

## Variance

For sensible measures of variance we really *need* some ground truthing data - pilot studies or measures from non-treatment populations

Finally, set a `seed` so that the generated random numbers are reproducible:
:::

```{r}
set.seed(38596)

# Sample size and factors
n <- 100
treatment <- factor(rep(c("A", "B"), each = n/2))      # 50 A, 50 B
sex       <- factor(rep(c("Male","Female"), times = n/2))

# --- Specify the true model on the MEAN scale ---
# Using default treatment (dummy) coding with baseline = treatment A, sex Male
beta_0      <- 400   # mean for (A, Male)
beta_trtB   <- -20    # difference (B - A) among Males when no interaction in the true DGP
beta_male <-  -200    # difference (Female - Male) in A when no interaction
beta_int    <-  0    # interaction: extra effect for (B,Female) beyond main effects

# Build the model matrix for ~ treatment * sex (includes intercept, both mains, and interaction)
X <- model.matrix(~ treatment + sex)

# Pack coefficients in the order of colnames(X)
coef_true <- c(beta_0, beta_trtB, beta_male)
names(coef_true) <- colnames(X)

# Mean for each observation implied by predictors
mu <- as.vector(X %*% coef_true)

# --- Residual SD goes here (this is where you include sd) ---
sigma <- 50
body_mass <- rnorm(n, mean = mu, sd = sigma)

dat <- data.frame(body_mass, treatment, sex)

# Fit models
m_main <- lm(body_mass ~ treatment + sex, data = dat)   # main effects only

```

```{r}

saveRDS(dat, file = "body_mass.RDS")

```


Let’s plot our simulated data:

```{r}
ggplot(dat, aes(x=treatment, 
               y=body_mass, 
               fill = sex)) + 
  gghalves::geom_half_violin(aes(split = treatment)) + 
  geom_boxplot(width = .1, 
               outlier.shape = NA,
               show.legend = FALSE) 
 # gghalves::geom_half_point(side = "l", range_scale = .4, 
   #     alpha = .5)

```



Let’s assume that we collected and analyzed this specific sample – what would have been the results?

```{r}
summary(m_main)
```

As you can see, in this simulated sample (based on a specific seed), the treatment effect is marginally significant (p < .05). But how likely are we to detect an effect with a sample of this size?


## Doing the power analysis

Now we need to repeatedly draw many samples and see how many of the analyses would have detected the existing effect. To do this, we put the code from above into a function called `sim`. We coded the function to either return the focal p-value (as default) or to print a model summary (helpful for debugging and testing the function). This function takes two parameters:

- `n` defines the required sample size
treatment_effect defines the treatment effect in the raw scale (i.e., reduction in body mass)

- We then use the `replicate` function to repeatedly call the `sim` function for 1000 iterations.


```{r}
set.seed(38596)

iterations <- 1000 # the number of Monte Carlo repetitions
n <- 100 # the size of our simulated sample

sim1 <- function(n=100, treatment_effect=-20, print=FALSE) {
# Sample size and factors
n <- n
treatment <- factor(rep(c("A", "B"), each = n/2))      # 50 A, 50 B
sex       <- factor(rep(c("Male","Female"), times = n/2))

# --- Specify the true model on the MEAN scale ---
# Using default treatment (dummy) coding with baseline = treatment A, sex Male
beta_0      <- 400   
beta_trtB   <- treatment_effect    # argument of interest
beta_male <-  -200    
beta_int    <-  0    

# Build the model matrix for ~ treatment * sex (includes intercept, both mains, and interaction)
X <- model.matrix(~ treatment + sex)

# Pack coefficients in the order of colnames(X)
coef_true <- c(beta_0, beta_trtB, beta_male)
names(coef_true) <- colnames(X)

# Mean for each observation implied by predictors
mu <- as.vector(X %*% coef_true)

# --- Residual SD goes here (this is where you include sd) ---
sigma <- 50
body_mass <- rnorm(n, mean = mu, sd = sigma)

dat <- data.frame(body_mass, treatment, sex)

# Fit models
m_main <- lm(body_mass ~ treatment + sex, data = dat)   # main effects only
  
  # this lm() call should be exactly the function that you use
  # to analyse your real data set
  p_value <- summary(m_main)$coefficients["treatmentB", "Pr(>|t|)"]
  
  if (print==TRUE) print(summary(res))
  else return(p_value)
}

# now run the sim() function a 1000 times and store the p-values in a vector:
p_values <- replicate(iterations, sim1(n=100))
```

How many of our 1000 virtual samples would have found the effect?

```{r}
table(p_values < .05)
```

Only **52%** of samples with the same size of n=100 result in a significant p-value.

## Sample size planning

Now we know that a sample size of 100 does not lead to a sufficient power. But what sample size would we need to achieve a power of at least 80%? In the simulation approach you need to test different n until you find the necessary sample size. We do this by wrapping the simulation code into a loop that continuously increases the n. We then store the computed power for each n.

```{r}
set.seed(38596)

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(100, 300, by=20) # test ns between 100 and 300


result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})
  
  # show the result after each run (not shown here in the tutorial)
  print(result)


```

Let's plot the result:

```{r}
ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()
```

Hence, with n = 200, (100 in each group), we have an 80% chance to detect the effect.

## Safe power analysis

As sensitivity analysis, we will apply a safeguard power analysis ([Perugini et al., 2014](https://doi.org/10.1177/1745691614528519)) that aims for the lower end of a two-sided 60% CI around the parameter of the treatment effect (the intercept is irrelevant). (Of course you can use any other value than 60%, but this is the value (tentatively) mentioned by the inventors of the safeguard power analysis.)

```{r}
beta_hat <- -21.819      # estimate
se <- 10.607            # standard error from model

z60 <- qnorm(1 - (1 - 0.60)/2)

ci_60_lower <- beta_hat - z60 * se
ci_60_upper <- beta_hat + z60 * se

c(ci_60_lower, ci_60_upper)

```

The *conservative* treatment difference to plan for is -12.89

```{r}
set.seed(38596)

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(100, 500, by=20) # test ns between 100 and 300


result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -12.89))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})
  
ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()

```

With that more conservative effect size assumption, we would need around 450 individuals, i.e., 225 per group.


## Smallest Effect of Interest

Many methodologists argue that we should not power for the expected effect size, but rather for the smallest effect size of interest (SESOI). In this case, a non-significant result can be interpreted as “We accept the H0, and even if a real effect existed, it most likely is too small to be relevant”.

We have to decide for ourselves what we value as the smallest effect size of interest:

```{r}
set.seed(38596)

# define all predictor and simulation variables.
iterations <- 1000
ns <- seq(50, 300, by=20) # 


result <- map_dfr(ns, ~{
  p_values <- replicate(iterations, sim1(n = .x, treatment_effect = -40))
  tibble(
    n = .x,
    power = mean(p_values < 0.05)
  )
})
  
ggplot(result, 
       aes(x=n, y=power)) + 
  geom_point() + 
  geom_line()+
  theme_bw()

```

If we decide that we are happy ignoring any difference that has <10% effect on body mass across treatments, then we can produce a reasonably powered experiment with only 50 individuals (25 per treatment). 

1. Define the statistical model that you will use to analyze your data (in our case: a linear regression). This sometimes is called the data generating mechanism or the data generating model.

2. Find empirical information about the parameters in the model. These are typically mean values, variances, group differences, regression coefficients, or correlations (in our case: a mean value (before treatment), a group difference, and an error variance). These numbers define the assumed state of the population.

3. Program your regression equation, which consists of a systematic (i.e., deterministic) part and (one or more) random error terms.

4. Do the simulation:

    a. Repeatedly sample data from your equation (only the random error changes in each run, the deterministic part stays the same).

    b. Compute the index of interest (typically: the p-value of a focal effect).

    c. Count how many samples would have detected the effect (i.e., the statistical power).

    d. Tune your simulation parameters until the desired level of power is achieved.
    
    Typically we would aim to change experimental sample numbers until we reach desired power, but we could also imagine making changes to measurement accuracy, or perhaps insect rearing parameters that will reduce variance?
