# Summarise {#sec-summarise}

```{r, echo = F, warning = F, message = F}
library(tidyverse)
library(janitor)
library(here)
source("R/booktem_setup.R")
source("R/my_setup.R")
penguins_raw <- read_csv(here("files", "penguins_raw.csv"))
```



::: {.callout-warning}

Re-open your `01_import_penguins_data.R` started in @sec-penguins and start to add these commands to your data importing and cleaning script:

:::


### Activity 1: Explore data structure

Before working with your data, it's essential to understand its underlying structure and content. In this section, we'll use powerful functions like `glimpse()`, `str()`, `summary()`, `head`()`, `tail()` and the add-on function `skimr::skim()` to thoroughly examine your dataset. These tools provide insights into data types, variable distributions, and sample records, helping you identify initial issues such as missing values or inconsistent data types. By gaining a clear understanding of your data's structure, you'll be better equipped to address any problems and proceed confidently with data cleaning and analysis.

When we run `glimpse()` we get several lines of output. The number of observations "rows", the number of variables "columns". Check this against the csv file you have - they should be the same. In the next lines we see variable names and the type of data. 


```{r, eval = T}

glimpse(penguins_raw)

```

We can see a dataset with 345 rows (including the headers) and 17 variables
It also provides information on the *type* of data in each column

* `<chr>` - means character or text data

* `<dbl>` - means numerical data


When we run `summary()` we get similar information, in addition for any numerical values we get summary statistics such as mean, median, min, max, quartile ranges and any missing (`NA`) values

```{r, eval = T}

summary(penguins_raw)

```

Finally the add-on package `skimr` provides the function `skimr::skim()` provides an easy to view set of summaries including column types, completion rate, number of unique variables in each column and similar statistical summaries along with a small histogram for each numeric variable. 

```{r}
library(skimr)
skim(penguins_raw)

```


**Q Based on our summary functions are any variables assigned to the wrong data type (should be character when numeric or vice versa)?**

`r mcq(c("Yes", answer = "No"))`


`r hide("Explanation")`

Although some columns like date might not be correctly treated as character variables, they are not strictly numeric either, all other columns appear correct

`r unhide()`


**Q Based on our summary functions do we have complete data for all variables?**

`r mcq(c("Yes", answer = "No"))`


`r hide("Explanation")`

No, they are 2 missing data points for body measurements (culmen, flipper, body mass), 11 missing data points for sex, 13/14 missing data points for blood isotopes (Delta N/C) and 290 missing data points for comments

`r unhide()`


